---
title: "Gaussian Copula Demo"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---


<style type="text/css">
.main-container {
  max-width: 1000px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, cache = T)
```

## Package Import

```{r}
library(tram)
library(mvtnorm)


```

## Wrapper Function 

```{r}

fit_mmlt_flexible <- function(dataframe,
                              seed=490,
                              marginal_formula="~ 1",
                              multivariate_formula="~ 1",
                              continuous_vars = c(),
                              discrete_vars = c(),
                              exclude_vars = c(), # removes a variable for the responses
                              continuous_formula = "default",
                              order=6
                              ){
  require(tram)
  require(qrng)
  set.seed(seed)
  
  # Check if all continuous and discrete variables are in the dataframe
  all_vars <- c(continuous_vars, discrete_vars)
  missing_vars <- setdiff(all_vars, names(dataframe))
  if (length(missing_vars) > 0) {
    warning(paste("Missing variables in dataframe:", paste(missing_vars, collapse = ", "), ", omitting"))
  }
  
  # make sure to exclude the marginal covariate from the responses, a regression on itself causes errors
  exclusion_variable <- ifelse(trimws(sub("~", "", multivariate_formula)) != "1", trimws(sub("~", "", multivariate_formula)), "")
  nm <- setdiff(names(dataframe), exclusion_variable) # if the argument is 1, the exclusion will be "", which is ignored on this line
  nm <- setdiff(nm, exclude_vars) 
  
  
  # Specify Optmizers
  op <- mltoptim()[1:3] # Order: auglag, spg, nloptr 
  
  m <- list() # Initialize m as a list of marginal models for the joint call below
  for (i in seq_along(nm)) {
    name <- nm[i]
    
    if(!(name %in% exclude_vars)){
      if (name %in% discrete_vars) {
        
        formula_string <- paste("R(", nm[i], ", as.R.interval= TRUE) ", marginal_formula)
        formula <- as.formula(formula_string)
        
        m[[i]] <- BoxCox(formula, data = dataframe, order=order)
        
      } else {
        if(!(name %in% continuous_vars )){
          warning(paste(name,"not categorized, including it as continuous"))
        }
        
        if(continuous_formula=="interval") {
          # The likelihood is  calculated over intervals
          # trans.fun. is non parametric likelihood
          
          formula_string <- paste("R(", nm[i], ", as.R.interval= TRUE) ", marginal_formula)
          formula <- as.formula(formula_string)
          
        } else if(continuous_formula=="ordered") {
          formula_string <- paste("R(", nm[i], ", as.R.ordered= TRUE) ", marginal_formula)
          formula <- as.formula(formula_string)
          
        } else {
          # log density
          # The likelihood is the continuous density on Bernstein Polynomial
          
          formula_string <- paste("R(", nm[i], ") ", marginal_formula)
          formula <- as.formula(formula_string)
        }
        
        m[[i]] <- BoxCox(formula, data = dataframe, order=order) # add the formula from above
        
      } 
    }
    
  }
  
  
  m$data <- dataframe
  m$formula <- as.formula(multivariate_formula)
  m$optim <- op
  
  # Always Joint here:
  m$domargins <- TRUE
  m$theta <- coef(m) # Use hot start
  mm <- do.call("mmlt", m)
  
  log_likelihood <- logLik(mm)

  
  n_par_total <- length(coef(mm))
  n_par_marginal <- sum(sapply(coef(mm, type = "marginal"), length)) 
  n_par_multivariate <- n_par_total - n_par_marginal
  n_obs <- nrow(dataframe)

  AIC <- 2*n_par_total - 2*as.numeric(log_likelihood)
  BIC <- n_par_total*log(n_obs) - 2*as.numeric(log_likelihood)
  
  coef_corr <- tryCatch( # If we have stratified correlation matrices, this will fail, only useful for combined correlations
    expr = {
        as.array(coef(mm, type="Corr"))[,,1]
    },error = function(e){  })
  
  coef_lambda <- coef(mm)
  standard_errors_lambda <- sqrt(diag(vcov(mm)))
  p_values <- 2 * pnorm(-abs( coef_lambda/ standard_errors_lambda))
  
  return(
    list(
      object = mm,
      coef_corr =  coef_corr,
      coef_lambda = coef_lambda,
      standard_errors_lambda = standard_errors_lambda,
      p_values = p_values,
      logLik = log_likelihood,
      AIC = AIC,
      BIC = BIC,
      n_par_total = n_par_total,
      n_par_marginal = n_par_marginal,
      n_par_multivariate = n_par_multivariate,
      n_obs = n_obs,
      n_var = length(nm)
    )
  )
}


```

## Example Dataset: Iris

```{r}
dataframe <- iris

library(GGally)
ggpairs(dataframe)
```



```{r}
str(dataframe)

```

## Different Copula Versions

Omitting the species:

```{r}
m1 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Species"),
                        order = 6
                        )
m1$coef_corr
m1$coef_lambda
m1$standard_errors_lambda
m1$p_values
m1$AIC
m1$BIC
m1$n_par_marginal
m1$n_par_multivariate
m1$n_par_total
m1$n_obs
m1$n_var

# if we only want to see the correlation p-values and format the nicely:
library(biostatUZH) # installation: devtools::install_github(repo = "felix-hof/biostatUZH")
correlation_pvals <- m1$p_values[m1$n_par_marginal+1:m1$n_par_multivariate]
cbind(names(correlation_pvals), format.pval(correlation_pvals))
```

Including Species as a discrete covariate:

```{r}
m2 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c("Species")
)

m2$coef_corr
m2$coef_lambda
m2$p_values
m2$AIC
m2$BIC
m2$n_par_marginal
m2$n_par_multivariate

```


Calculating a correlation matrix per species:

```{r}
m3 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c()
)

m3$coef_corr
m3$coef_lambda
m3$p_values
m3$AIC
m3$BIC
m3$n_par_marginal
m3$n_par_multivariate

```

Correlation matrix per species, but the transformation is limited to a polynomial to order of 2:

```{r}
m4 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        order=2
)

m4$coef_corr
m4$coef_lambda
m4$p_values
m4$AIC
m4$BIC
m4$n_par_marginal
m4$n_par_multivariate

```

Including a marginal species covariate, and a correlation matrix per species:

```{r}
m5 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ Species",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c()
)

m5$coef_corr
m5$coef_lambda
m5$p_values
m5$AIC
m5$BIC
m5$n_par_marginal
m5$n_par_multivariate


```

Adding a scale term to the marginal models to account for different dispersions:


```{r}
m6 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ Species | Species",
                        multivariate_formula="~ 1 ",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c()
)

m6$coef_corr
m6$coef_lambda
m6$p_values
m6$AIC
m6$BIC
m6$n_par_marginal
m6$n_par_multivariate

```

Adding a conditional baseline transformation that depends on the species as well as a correlation matrix per species:

```{r}
m7 <- fit_mmlt_flexible(dataframe,
                        marginal_formula=" | Species ~ 1",
                        multivariate_formula="~ Species ",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Species")
)

m7$coef_corr
m7$coef_lambda
m7$p_values
m7$AIC
m7$BIC
m7$n_par_marginal
m7$n_par_multivariate

```

Comparing the AIC and BIC of all these models:


```{r}
c(m1$AIC, m2$AIC, m3$AIC, m4$AIC, m5$AIC, m6$AIC, m7$AIC)
c(m1$BIC, m2$BIC, m3$BIC, m4$BIC, m5$BIC, m6$BIC, m7$BIC)

```

## Visualisation of the correlation matrix

```{r}
library(pheatmap)
library(viridis)

pheatmap(m2$coef_corr,
         color = viridis(1000),
         clustering_method = "ward.D2",
         display_numbers = TRUE)  


library(corrplot)
corrplot(m2$coef_corr, method = "square", 
         type = "lower", 
         order = "hclust", 
         hclust.method = "ward.D2", 
         tl.pos = "ld",
         tl.cex = 0.5, 
         tl.col = "black", 
         tl.srt = 45,
         addCoef.col = "black") 

library(ggcorrplot)
ggcorrplot(m2$coef_corr, lab = TRUE) + 
  scale_fill_gradientn(colors = viridis(1000, option = 'D'))
```

## Likelihood Ratio Test for Model Comparison

We can use the likelihood arguments from inside the model together with the known number of parameters of each.
*If* the two models are nested (i.e. one contains all parameters of the other), the likelihood ratio follows - asymptotically - a Chi Square distribution with the degrees of freedom being the difference between these two nested models.
The resulting p-value can be used to quantify whether the larger model improves the model fit to the data.

```{r}
lr_test_copula <- function(outer_model, inner_model){
  
  if(outer_model$n_par_total < inner_model$n_par_total){ # ensure correct order
    temp <- outer_model
    outer_model <- inner_model
    inner_model <- temp
  }
  
  if (any(!names(inner_model$coef_lambda) %in% names(outer_model$coef_lambda))) {   # Check if any coefficients are missing
    missing_coefs <- names(inner_model$coef_lambda)[!names(inner_model$coef_lambda) %in% names(outer_model$coef_lambda)]
    print(
      warning("The following coefficients from the inner model are not present in the outer model: ", paste(missing_coefs, collapse = ", "))
    )  
    return(NULL)
  }

  
  outer_log_likelihood <- as.numeric(outer_model$logLik)
  outer_n_par <- outer_model$n_par_total
  
  inner_log_likelihood <- as.numeric(inner_model$logLik)
  inner_n_par <- inner_model$n_par_total
  
  # Pvalue
  lr_statistic <- -2*(inner_log_likelihood - outer_log_likelihood)
  delta_df <- outer_n_par-inner_n_par
  p_value <- pchisq(lr_statistic, df=delta_df, lower.tail=FALSE)

  result_table <- data.frame(
    Model = c("Inner Model", "Outer Model"),
    npar = c(inner_n_par, outer_n_par),
    AIC = c(inner_model$AIC, outer_model$AIC),
    BIC = c(inner_model$BIC, outer_model$BIC),
    logLik = c(inner_log_likelihood, outer_log_likelihood),
    deviance = c(-2 * inner_log_likelihood, -2 * outer_log_likelihood),
    Chisq = c(NA, lr_statistic),
    Df = c(NA, delta_df),
    "Pr(>Chisq)" = c(NA, p_value)
  )
  
  print(result_table, row.names = FALSE)

  return(p_value)
}

```





```{r}
m_outer <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        order=6
)

m_inner <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        order=2
)

lr_test_copula(m_outer, m_inner)


m_inner <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        order=5
)

lr_test_copula(m_outer, m_inner)

```


```{r}
m_outer <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c()
)

m_inner <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Species")
)

lr_test_copula(m_outer, m_inner)


```


```{r}
 m_inner <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Petal.Width", "Species")
)

m_outer <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Species")
)

lr_test_copula(m_outer, m_inner)


```


```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```

## Session Info

```{r, echo=F}
date()

sessionInfo()

# CPU and RAM info if benchmarkme is installed
if(require(benchmarkme)) cat(sprintf("\nCPU: %s | Cores: %d\nRAM: %.1f GiB\n", get_cpu()$model_name, get_cpu()$no_of_cores, as.numeric(get_ram()) / 1024^3))

```

