---
title: "Gaussian Copula Demo"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, cache = T)
```

## Package Import

```{r}
library(tram)
library(mvtnorm)


```

## Wrapper Function 

```{r}

fit_mmlt_flexible <- function(dataframe,
                              seed=490,
                              marginal_formula="~ 1",
                              multivariate_formula="~ 1",
                              continuous_vars = c(),
                              discrete_vars = c(),
                              exclude_vars = c(), # removes a variable for the responses
                              continuous_formula = "default",
                              order=6
                              ){
  require(tram)
  require(qrng)
  set.seed(seed)
  
  # Check if all continuous and discrete variables are in the dataframe
  all_vars <- c(continuous_vars, discrete_vars)
  missing_vars <- setdiff(all_vars, names(dataframe))
  if (length(missing_vars) > 0) {
    warning(paste("Missing variables in dataframe:", paste(missing_vars, collapse = ", "), ", omitting"))
  }
  
  # make sure to exclude the marginal covariate from the responses, a regression on itself causes errors
  exclusion_variable <- ifelse(trimws(sub("~", "", multivariate_formula)) != "1", trimws(sub("~", "", multivariate_formula)), "")
  nm <- setdiff(names(dataframe), exclusion_variable) # if the argument is 1, the exclusion will be "", which is ignored on this line
  nm <- setdiff(nm, exclude_vars) # if the argument is 1, the exclusion will be "", which is ignored on this line
  
  
  # Specify Optmizers
  op <- mltoptim()[1:3] # Order: auglag, spg, nloptr 
  
  m <- list() # Initialize m as a list of marginal models for the joint call below
  for (i in seq_along(nm)) {
    name <- nm[i]
    
    if(!(name %in% exclude_vars)){
      if (name %in% discrete_vars) {
        
        formula_string <- paste("R(", nm[i], ", as.R.interval= TRUE) ", marginal_formula)
        formula <- as.formula(formula_string)
        
        m[[i]] <- BoxCox(formula, data = dataframe, order=order)
        
      } else {
        if(!(name %in% continuous_vars )){
          warning(paste(name,"not categorized, including it as continuous"))
        }
        
        if(continuous_formula=="interval") {
          # The likelihood is  calculated over intervals
          # trans.fun. is non parametric likelihood
          
          formula_string <- paste("R(", nm[i], ", as.R.interval= TRUE) ", marginal_formula)
          formula <- as.formula(formula_string)
          
        } else if(continuous_formula=="ordered") {
          formula_string <- paste("R(", nm[i], ", as.R.ordered= TRUE) ", marginal_formula)
          formula <- as.formula(formula_string)
          
        } else {
          # log density
          # The likelihood is the continuous density on Bernstein Polynomial
          
          formula_string <- paste("R(", nm[i], ") ", marginal_formula)
          formula <- as.formula(formula_string)
        }
        
        m[[i]] <- BoxCox(formula, data = dataframe, order=order) # add the formula from above
        
      } 
    }
    
  }
  
  
  m$data <- dataframe
  m$formula <- as.formula(multivariate_formula)
  m$optim <- op
  
  # Always Joint here:
  m$domargins <- TRUE
  m$theta <- coef(m) # Use hot start
  mm <- do.call("mmlt", m)
  
  log_likelihood <- logLik(mm)
  standard_errors <- sqrt(diag(vcov(mm)))
  
  AIC <- 2*length(coef(mm)) - 2*as.numeric(log_likelihood)
  BIC <- length(coef(mm))*log(nrow(dataframe)) - 2*as.numeric(log_likelihood)
  
  return(
    list(
      object = mm,
      coefficients = coef(mm),
      standard_errors = standard_errors,
      logLik = log_likelihood,
      AIC = AIC,
      BIC = BIC
    )
  )
}

```

## Example Dataset: Iris

```{r}
dataframe <- iris

library(GGally)
ggpairs(dataframe)
```



```{r}
str(dataframe)

```

## Different Copula Versions

Omitting the species:

```{r}
m1 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Species"),
                        order = 6
                        )
coef(m1)
m1$AIC
m1$BIC

```

Including Species as a discrete covariate:

```{r}
m2 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ 1",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c("Species")
)
coef(m2)
m2$AIC
m2$BIC

```


Calculating a correlation matrix per species:

```{r}
m3 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c()
)
coef(m3)
m3$AIC
m3$BIC

```

Correlation matrix per species, but the transformation is limited to a polynomial to order of 2:

```{r}
m4 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ 1",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        order=2
)
coef(m4)
m4$AIC
m4$BIC

```

Including a marginal species covariate, and a correlation matrix per species:

```{r}
m5 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ Species",
                        multivariate_formula="~ Species",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c()
)
coef(m5)
m5$AIC
m5$BIC


```

Adding a scale term to the marginal models to account for different dispersions:


```{r}
m6 <- fit_mmlt_flexible(dataframe,
                        marginal_formula="~ Species | Species",
                        multivariate_formula="~ 1 ",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c()
)
coef(m6)
m6$AIC
m6$BIC

```

Adding a conditional baseline transformation that depends on the species as well as a correlation matrix per species:

```{r}
m7 <- fit_mmlt_flexible(dataframe,
                        marginal_formula=" | Species ~ 1",
                        multivariate_formula="~ Species ",
                        continuous_vars = c("Sepal.Length", "Sepal.Width",  "Petal.Length", "Petal.Width" ),
                        discrete_vars = c(),
                        exclude_vars = c("Species")
)
coef(m7)
m7$AIC
m7$BIC

```

Comparing the AIC and BIC of all these models:


```{r}
c(m1$AIC, m2$AIC, m3$AIC, m4$AIC, m5$AIC, m6$AIC, m7$AIC)
c(m1$BIC, m2$BIC, m3$BIC, m4$BIC, m5$BIC, m6$BIC, m7$BIC)

```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```

## Session Info

```{r, echo=F}
date()

sessionInfo()

# CPU and RAM info if benchmarkme is installed
if(require(benchmarkme)) cat(sprintf("\nCPU: %s | Cores: %d\nRAM: %.1f GiB\n", get_cpu()$model_name, get_cpu()$no_of_cores, as.numeric(get_ram()) / 1024^3))

```

